Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-04 appendix.tex
restructured

\subsection{Thin-disk limit}
For thin disks ($\epsilon\ll1$), in which 
$\ln\rho\simeq -\zhat^2/2$ and $f(\zhat)\simeq \zhat$, 
% by expanding the background density and vertical shear profile in powers
% of $z/r$. To lowest order we obtain 
% \begin{align}
%   &\frac{\rho}{\rho_0} \simeq 
%   \exp{\left(-\frac{\hat{z}^2}{2}\right)} \equiv w(\hat{z}),\label{thin_dens}\\
%   &\frac{d\Omega^2}{d\hat{z}} \simeq \epsilon^2q\Omega_k^2\hat{z}, \label{thin_vshear}
% \end{align} 
%With this approximation and those in \S\ref{integral_relation}, 
Eq. \ref{iso_ode1} becomes  
\begin{align}\label{iso_ode3}
  W^{\prime\prime} - \left(1 + \ii q\epsilon
    \hat{k}\right)\hat{z}W^\prime  +
  \hat{\sigma}^2\left(1+\hat{k}^2\right)W = 
  0.
\end{align}
We remark that for large $|\hat{k}|$, Eq. \ref{iso_ode3} is the same as
that derived by \cite{nelson13}, although we have taken a different
route.

To complete the problem we must specify boundary conditions. The 
maximum vertical domain size should be limited by the thin
disk and low-frequency approximations (i.e. $z\lesssim r$).     
It is, however, common to take $\mathrm{max}|z|\to\infty$ and impose
that the kinetic  energy remain bounded at infinity. This 
is appropriate for the density field because both the true density profile 
and its thin-disk approximation decay rapidly away from the midplane.

@@

\subsubsection{Stability in the absence of vertical shear}\label{iso_stable}
When $q=0$, Eq. \ref{iso_ode3} is
\begin{align}\label{hermite_ode}
  \left[w(\hat{z})W^\prime \right]^\prime + nW
  w(\hat{z}) =0, 
\end{align}
where $w(\zhat)=\exp{(-\zhat^2/2)}$ and we have defined a new eigenvalue 
\begin{align}
  n \equiv \hat{\sigma}^2(1+\hat{k}^2). 
\end{align} 
Eq. \ref{hermite_ode} is Hermite's differential equation. If we impose
that the kinetic energy density remain bounded at infinity, then the
solutions are Hermite polynomials  
\begin{align}
  W \propto \He_n(\hat{z}),
\end{align}
and $n$ is a non-negative integer. This translates to a real
eigenfrequency $\sigma$ such that
\begin{align}
  \left|\hat{\sigma}\right| = \sqrt{n}
  \left(1+\hat{k}^2\right)^{-1/2}. \label{iso_stable_disp}
\end{align}
Since we have assumed $|\sigma^2|\ll \kappa^2\sim \Omega_k^2$, our
analysis is only valid for large wavenumbers such that $\hat{k}^2\gg   
n-1$. Physically, this corresponds to radial length-scales much
smaller than the local disk scale height. Note that
Eq. \ref{iso_stable_disp} is implied by Eq. 54 in
\cite{lubow93} after setting $\gamma=1$ and making the low-frequency,
Keplerian disk approximation in that equation.

@@

\subsubsection{Polynomial solutions with vertical
  shear}\label{iso_poly}
More generally, we can seek power-series solution to
Eq. \ref{iso_ode3}, 
\begin{align}
  W(\zhat) = \sum_{l=0}^\infty a_l\zhat^l. 
\end{align}
Then the coefficients must satisfy the recurrence relation
\begin{align}
  (l+2)(l+1)a_{l+2} +
  \left[n - l\left(1+\ii \epsilon q  \hat{k}\right)\right] a_l = 0, 
\end{align}
where $n$ is not necessarily an integer. Indeed, if we demand
a polynomial of degree $L$ as the solution, then the eigenfrequency
must satisfy
\begin{align}\label{sig2_iso}
  \hat{\sigma}^2 = L\left(\frac{1+\ii q \epsilon
      \hat{k}}{1+\hat{k}^2}\right).
\end{align}
The eigenfrequency $\hat{\sigma}$ is therefore complex for
$L\geq1$. The dispersion relation Eq. \ref{sig2_iso} can also be
derived by setting $\beta=0$ in the more general dispersion relation 
Eq. \ref{relax_disp} with coefficients given in Appendix
\ref{relax_coeff}. We are interested in 
unstable modes for which $\hat{\nu}>0$ and is given by
\begin{align}\label{simple_growth}
  \hat{\nu} =\sqrt{
    \frac{L}{2\left(1+\hat{k}^2\right)}\left(\sqrt{1+q^2\epsilon^2\hat{k}^2} - 
      1\right)}. 
\end{align}
For fixed $q$ and $\epsilon$, the growth rate vanishes for both
$\hat{k}^2\to0$ and $\hat{k}^2\to\infty$. 
The maximum growth rate
occurs at the optimum wavenumber $\hat{k}_\mathrm{opt}$
\begin{align}
  |\hat{k}_\mathrm{opt}| = \sqrt{\frac{2+|\epsilon q|}{|\epsilon q|}},
\end{align}
and
\begin{align}
  \mathrm{max}\left(\hat{\nu}\right) =\frac{\sqrt{L}|\epsilon
    q|}{2\sqrt{1+|\epsilon q|}}. \label{iso_max_growth}
\end{align}
Since $L\geq1$, there is no limit to the maximum growth rate. This is
an artifact of the thin-disk approximation because the vertical 
shear $d\Omega^2/dz\propto z$ increases indefinitely with 
height. However, Eq. \ref{iso_max_growth} compares well with
numerical solutions for small $L$ (see \S\ref{vertiso_pertiso}).

>>>>

\subsection{Thin-disk limit}
For thin disks ($\epsilon\ll1$), in which 
$\ln\rho\simeq -\zhat^2/2$ and $f(\zhat)\simeq \zhat$, 
% by expanding the background density and vertical shear profile in powers
% of $z/r$. To lowest order we obtain 
% \begin{align}
%   &\frac{\rho}{\rho_0} \simeq 
%   \exp{\left(-\frac{\hat{z}^2}{2}\right)} \equiv w(\hat{z}),\label{thin_dens}\\
%   &\frac{d\Omega^2}{d\hat{z}} \simeq \epsilon^2q\Omega_k^2\hat{z}, \label{thin_vshear}
% \end{align} 
%With this approximation and those in \S\ref{integral_relation}, 
Eq. \ref{iso_ode1} becomes  
\begin{align}\label{iso_ode3}
  W^{\prime\prime} - \left(1 + \ii q\epsilon
    \hat{k}\right)\hat{z}W^\prime  +
  \hat{\sigma}^2\left(1+\hat{k}^2\right)W = 
  0.
\end{align}
We remark that for large $|\hat{k}|$, Eq. \ref{iso_ode3} is the same as
that derived by \cite{nelson13}, although we have taken a different
route.  
% To complete the problem we must specify boundary conditions. The 
% maximum vertical domain size should be limited by the thin
% disk and low-frequency approximations (i.e. $z\lesssim r$).     
%It is, however, common to take $\mathrm{max}|z|\to\infty$ and 
We impose the kinetic energy remain bounded at infinity. Even though
we have adopted the thin-disk approximation, an infinite
disk is appropriate in the sense that both the true density profile 
and its thin-disk approximation decay rapidly away from the midplane.

@@

\subsubsection{Stability in the absence of vertical shear}\label{iso_stable}
When $q=0$, Eq. \ref{iso_ode3} is
\begin{align}\label{hermite_ode}
  \left[w(\hat{z})W^\prime \right]^\prime + nW
  w(\hat{z}) =0, 
\end{align}
where $w(\zhat)=\exp{(-\zhat^2/2)}$ and we have defined a new eigenvalue 
\begin{align}
  n \equiv \hat{\sigma}^2(1+\hat{k}^2). 
\end{align} 
Eq. \ref{hermite_ode} is Hermite's differential equation. If we impose
that the kinetic energy density remain bounded at infinity, then the
solutions are Hermite polynomials  
\begin{align}
  W \propto \He_n(\hat{z}),
\end{align}
and $n$ is a non-negative integer. This translates to a real
eigenfrequency $\sigma$ such that
\begin{align}
  \left|\hat{\sigma}\right| = \sqrt{n}
  \left(1+\hat{k}^2\right)^{-1/2}. \label{iso_stable_disp}
\end{align}
Since we have assumed $|\sigma^2|\ll \kappa^2\sim \Omega_k^2$, our
analysis is only valid for large wavenumbers such that $\hat{k}^2\gg   
n-1$. Physically, this corresponds to radial length-scales much
smaller than the local disk scale height. Note that
Eq. \ref{iso_stable_disp} is implied by Eq. 54 in
\cite{lubow93} after setting $\gamma=1$ and making the low-frequency,
nearly-Keplerian approximation in that equation.

@@

\subsubsection{Polynomial solutions with vertical
  shear}\label{iso_poly}
In the presence of vertical shear ($q\neq0$) we can still seek power-series solution to
Eq. \ref{iso_ode3}, 
\begin{align}
  W(\zhat) = \sum_{l=0}^\infty a_l\zhat^l. 
\end{align}
Then the coefficients must satisfy the recurrence relation
\begin{align}
  (l+2)(l+1)a_{l+2} +
  \left[n - l\left(1+\ii \epsilon q  \hat{k}\right)\right] a_l = 0, 
\end{align}
where $n$ is not necessarily an integer. Indeed, if we demand
a polynomial of degree $L$ as the solution, then the eigenfrequency
must satisfy
\begin{align}\label{sig2_iso}
  \hat{\sigma}^2 = L\left(\frac{1+\ii q \epsilon
      \hat{k}}{1+\hat{k}^2}\right).
\end{align}
The eigenfrequency $\hat{\sigma}$ is therefore complex for
$L\geq1$. The dispersion relation Eq. \ref{sig2_iso} can also be
derived by setting $\beta=0$ in the more general dispersion relation 
Eq. \ref{relax_disp} with coefficients given in Appendix
\ref{relax_coeff}. We are interested in 
unstable modes for which $\hat{\nu}>0$ and is given by
\begin{align}\label{simple_growth}
  \hat{\nu} =\sqrt{
    \frac{L}{2\left(1+\hat{k}^2\right)}\left(\sqrt{1+q^2\epsilon^2\hat{k}^2} - 
      1\right)}. 
\end{align}
For fixed $q$ and $\epsilon$, the growth rate vanishes for both
$\hat{k}^2\to0$ and $\hat{k}^2\to\infty$. 
The maximum growth rate
occurs at the optimum wavenumber $\hat{k}_\mathrm{opt}$
\begin{align}
  |\hat{k}_\mathrm{opt}| = \sqrt{\frac{2+|\epsilon q|}{|\epsilon q|}},
\end{align}
and
\begin{align}
  \mathrm{max}\left(\hat{\nu}\right) =\frac{\sqrt{L}|\epsilon
    q|}{2\sqrt{1+|\epsilon q|}}. \label{iso_max_growth}
\end{align}
Since $L\geq1$, there is no limit to the maximum growth rate. This is
an artifact of the thin-disk approximation because the magnitude of vertical 
shear $|d\Omega^2/dz|\propto |z|$ increases indefinitely with 
height. However, Eq. \ref{iso_max_growth} compares well with
numerical solutions for small $L$ (see \S\ref{vertiso_pertiso}).

### 2015-03-04 application.tex
restructured

\subsubsection{Comparing the VSI in the inner and outer disk} 
Here we compare the VSI at $r=50$AU and $r=5$AU. We first show in
Fig. \ref{beta_compare}  the thermal timescale $\beta$ for a
perturbation wavenumber $\khat=30$. For comparison we also plot
$\beta_\mathrm{crit}$ at these radii.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_50AU}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_5AU}
  \caption{Vertical velocity perturbations of the VSI in the MMSN at 
    $r=50$AU (top) and $r=5$AU (bottom). These are calculated using the height
    dependent thermal relaxation given by Eq. \ref{beta_mmsn_simp}. The real (imaginary) part of 
    $\delta v_z$ is shown as
    the solid (dotted) curves. The eigenfunction has been normalized
    such that $\imag(\delta v_z)=0$ at the vertical boundaries, and
    $\mathrm{max}|\delta v_z|=1$. \label{mmsn_eigenvz}}
\end{figure}

>>>>

\subsection{Comparing the VSI in the inner and outer parts of the MMSN} 
Here we compare the VSI in the MMSN at $r=50$AU and $r=5$AU. We first show in
Fig. \ref{beta_compare}  the thermal timescale $\beta$ for a
perturbation wavenumber $\khat=30$. For comparison we also plot
$\beta_\mathrm{crit}$ at these radii.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_50AU}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_5AU}
  \caption{Vertical velocity perturbations of the VSI in the MMSN at 
    $r=50$AU (top) and $r=5$AU (bottom). These are calculated using the height
    dependent thermal relaxation given by Eq. \ref{beta_mmsn_simp}. The real (imaginary) part of 
    $\delta v_z$ is shown as
    the solid (dotted) curves. % The eigenfunction has been normalized
    % such that $\imag(\delta v_z)=0$ at the vertical boundaries, and
    % $\mathrm{max}|\delta v_z|=1$. 
    \label{mmsn_eigenvz}}
\end{figure}

### 2015-03-04 linear_adia.tex
restructured

The thin disk approximation allows us to develop explicit solutions,
but some general results may be established without it (see,
e.g. Appendix \ref{global_corr} and \S\ref{iso_pert}).

\subsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
It is instructive to first ask  the more analytically tractable
question: how 
do the eigenfrequencies and eigenfunctions change when we change
$t_c=\beta\Omega_k^{-1}$ from zero to a small but finite value? For
sufficiently small $\beta$, for which the perturbations may be 
considered nearly-isothermal, we expect the solution to only
differ slightly from a case with $\beta\equiv 0$.

To determine the (assumed) small effect of introducing a short thermal
relaxation, we linearize Eq. \ref{nearly_iso_explicit} about a known 
solution for $\beta\equiv0$, which we choose below. Let  
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, \delta v_z\to \delta v_z+\delta
  v_{z1},\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align}
which implies 
\begin{align}
  \chi \to 1 + \delta\chi = 1 + \ii \hat{\sigma}\left(\gamma-1\right)\delta\beta.
\end{align}
Inserting Eq. \ref{nearly_iso_pert} into Eq. \ref{nearly_iso_explicit}
and keeping only first order terms, we obtain  
\begin{align}
 0 =& \delta v_{z1}^{\prime\prime} - z\left(1 + \ii\epsilon q
   k\right)\delta v_{z1}^\prime\notag\\
 &+ \left[2\hat{\sigma}\delta\hat{\sigma}\left(1 + \khat^2\right) +
   \left(\hat{\sigma}^2 - 1\right)\delta\chi \right.\notag\\
&\phantom{++} \left. + \zhat^2\left(\khat^2 -
     \ii\epsilon q \khat\right)\delta\chi\right]\delta v_z\notag\\
 &+\left[\hat{\sigma}^2\left(1 + \khat^2\right) - \left(1+\ii\epsilon
     q \khat\right)\right]\delta v_{z1}.\label{nearly_iso4}
\end{align}

In Eq. \ref{nearly_iso_pert} and \ref{nearly_iso4}, $\delta v_z$ is to
a solution to Eq. \ref{nearly_iso_explicit} with 
$\beta=0$. For definiteness, let us consider the simple solution 
\begin{align}
  \beta\equiv 0, \quad \delta v_z = 1,\quad \hat{\sigma}^2 = \frac{1 +
  \ii\epsilon q \hat{k}}{1+\hat{k}^2}, \label{pert_basic} 
\end{align}
which solves Eq. \ref{nearly_iso_explicit} because $\delta
v_z=$constant, $\chi=1$ so that $C=0$, and the eigenfrequency makes
$B=0$. Then we may seek 
\begin{align}
  \delta v_{z1} = a \zhat^2 + b,
\end{align}
where $a$, $b$ are constants.

% We could proceed by considering the entire set of polynomial solutions
% for $W$, and hence for $\delta v_z$, as derived in \S\ref{iso_poly}, but it
% is simplest to look at the fundamental mode.

Inserting the above expressions for $\delta v_z$, $\delta v_{z1}$ and
$\hat{\sigma}^2$ into Eq. \ref{nearly_iso4}, then balancing
coefficients of $\zhat^2$ and constants, we require
\begin{align}
  0 &= 2a + 2\hat{\sigma}\left(1+\khat^2\right)\delta\hat{\sigma} +
  \left(\hat{\sigma}^2 -1\right)\delta\chi,\notag\\
  0&= 2a\left(1+\ii\epsilon q \khat\right) - \left(\khat^2 - \ii\epsilon
    q \khat\right)\delta\chi.
\end{align}

@@

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
$\khat\gg1$, we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3.\label{relax_cond2}
\end{align}
This is a pair of simultaneous equations for
$\hat{\omega}$ and $\beta$.  
% Since our discussion of stable solutions without vertical shear yield
% a dispersion relation $\hat{\omega}\propto \khat^{-1}$ for $\khat\gg
% 1$ (\S\ref{iso_stable}, Eq. \ref{iso_stable_disp}).
We seek solutions to 
Eq. \ref{relax_cond1}---\ref{relax_cond2} with  
% we wish to seek a criterion independent of k 
% \begin{align}
%   \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
%\end{align}
\begin{align}
  \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
  |\hat{\omega}\khat|=O(1). 
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$. 
For a thin disk $\epsilon\ll 1$, we infer 
\begin{align}\label{iso_vsi_cond}
  \beta \lesssim \frac{|\epsilon q|}{\gamma-1} \equiv
  \beta_\mathrm{crit} 
\end{align}
is required for the fundamental VSI to operate. We will test this
dependence numerically in \S\ref{bcrit_num_test}. 
From our discussion in \S\ref{relax_pert} we expect   
$\beta<\beta_\mathrm{crit}$ is required for the fundamental mode to be
unstable. That is, when $\beta<\beta_\mathrm{crit}$, perturbations are 
effectively isothermal.

@@

\subsection{Isothermal perturbations}\label{iso_pert}

>>>>

% The thin disk approximation allows us to develop explicit solutions,
% but some general results may be established without it (see,
% e.g. Appendix \ref{global_corr} and \S\ref{iso_pert}).

\subsection{Isothermal perturbations}\label{iso_pert}
Isothermal perturbations correspond to $\beta = 0$ or $\chi = 1$. This
case has been investigated numerically by \cite{nelson13} and 
\cite{mcnally14} in the thin disk limit, the latter of which employed
the VGSB framework.

In Appendix \S\ref{iso_discuss}, we give an analytical description of the
linear problem with $\beta=0$. There, we establish more general
results that do not require the thin disk approximation, namely 
a necessary condition for instability and an upper bound to the
instability growth rate. We also develop analytic solutions
with and without vertical shear in the thin disk limit, including
explicit expressions for wave frequencies and growth rates, the
methods and results for which motivates the solution procedure
employed in the following sections.

\subsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
We are interested in the case $\beta\neq 0$, but it is instructive to
first ask  the more analytically tractable question: how 
do the eigenfrequencies and eigenfunctions change when we change
$t_c=\beta\Omega_k^{-1}$ from zero to a small but finite value? For 
sufficiently small $\beta$, for which the perturbations may be 
considered nearly-isothermal, we expect the solution to only
differ slightly from a case with $\beta\equiv 0$.

For definiteness, let us consider the simple solution 
\begin{align}
  \beta\equiv 0, \quad \delta v_z = 1,\quad \hat{\sigma}^2 = \frac{1 +
  \ii\epsilon q \hat{k}}{1+\hat{k}^2}, \label{pert_basic} 
\end{align}
which solves Eq. \ref{nearly_iso_explicit} since $\delta
v_z=$constant and $\chi=1$ so that $B=C=0$. 
This is the fundamental or `corrugation' mode and represents
the entire disk column moving vertically. These modes have been
observed to dominate numerical simulations \citep{nelson13,stoll14},
and are the ones we find to develop in protoplanetary disks
with a realistic estimate for thermal timescales
(\S\ref{application}).

To determine the (assumed) small effect of introducing a short thermal
relaxation, we linearize Eq. \ref{nearly_iso_explicit} about the above
solution for $\beta\equiv0$. We write 
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, \delta v_z\to \delta v_z+\delta
  v_{z1},\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align}
which implies 
\begin{align}
  \chi \to 1 + \delta\chi = 1 + \ii \hat{\sigma}\left(\gamma-1\right)\delta\beta,
\end{align}
with $\delta v_z$ and $\hat{\sigma}$ given by Eq. \ref{pert_basic}. We
may then seek  
\begin{align}
  \delta v_{z1} = a \zhat^2 + b,
\end{align}
where $a$, $b$ are constants.

Inserting Eq. \ref{nearly_iso_pert} into Eq. \ref{nearly_iso_explicit}
and keeping only first order terms yield the requirements, after
balancing constants and coefficients of $\zhat^2$, 
% we obtain  
% \begin{align}
%  0 =& \delta v_{z1}^{\prime\prime} - z\left(1 + \ii\epsilon q
%    k\right)\delta v_{z1}^\prime\notag\\
%  &+ \left[2\hat{\sigma}\delta\hat{\sigma}\left(1 + \khat^2\right) +
%    \left(\hat{\sigma}^2 - 1\right)\delta\chi \right.\notag\\
% &\phantom{++} \left. + \zhat^2\left(\khat^2 -
%      \ii\epsilon q \khat\right)\delta\chi\right]\delta v_z\notag\\
%  &+\left[\hat{\sigma}^2\left(1 + \khat^2\right) - \left(1+\ii\epsilon
%      q \khat\right)\right]\delta v_{z1}.\label{nearly_iso4}
% \end{align}
% In Eq. \ref{nearly_iso_pert} and \ref{nearly_iso4}, $\delta v_z$ is to
% a solution to Eq. \ref{nearly_iso_explicit} with 
% $\beta=0$. 
% We could proceed by considering the entire set of polynomial solutions
% for $W$, and hence for $\delta v_z$, as derived in \S\ref{iso_poly}, but it
% is simplest to look at the fundamental mode. 
%Inserting the above expressions for $\delta v_z$, $\delta v_{z1}$ and
%$\hat{\sigma}^2$ into Eq. \ref{nearly_iso4}, then balancing
%coefficients of $\zhat^2$ and constants, we require
\begin{align}
  0 &= 2a + 2\hat{\sigma}\left(1+\khat^2\right)\delta\hat{\sigma} +
  \left(\hat{\sigma}^2 -1\right)\delta\chi,\notag\\
  0&= 2a\left(1+\ii\epsilon q \khat\right) - \left(\khat^2 - \ii\epsilon
    q \khat\right)\delta\chi.
\end{align}

@@

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
$\khat\gg1$, we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3.\label{relax_cond2}
\end{align}
This is a pair of simultaneous equations for
$\hat{\omega}$ and $\beta$. In Appendix \ref {iso_stable} we discuss
stable solutions without vertical shear which yield
$\hat{\omega}\propto \khat^{-1}$ for $\khat\gg 1$
(Eq. \ref{iso_stable_disp}),  we are thus motivated to seek solutions
to Eq. \ref{relax_cond1}---\ref{relax_cond2} with  
% we wish to seek a criterion independent of k 
% \begin{align}
%   \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
%\end{align}
\begin{align}
  \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
  |\hat{\omega}\khat|=O(1). 
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}\label{beta_crit0}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$.

From our discussion in \S\ref{relax_pert}, which indicate introducing
a finite thermal relaxation is stabilizing, we expect 
Eq. \ref{beta_crit0} is an upper limit for instability. For a thin
disk $\epsilon\ll 1$, we thus infer 
\begin{align}\label{iso_vsi_cond}
  \beta \lesssim \frac{|\epsilon q|}{\gamma-1} \equiv
  \beta_\mathrm{crit} 
\end{align}
is required for the fundamental VSI to operate. That is, when
$\beta<\beta_\mathrm{crit}$, perturbations are  effectively isothermal. 
We will test Eq. \ref{iso_vsi_cond} numerically in
\S\ref{bcrit_num_test}.

@@

### 2015-03-04 summary.tex
restructured

We included a source term in the energy equation which relaxes the disk
temperature to its initial value on a timescale
$t_c=\beta\Omega_k^{-1}$. This approach to model thermal relaxation was also
employed in the global numerical simulations of \cite{nelson13}, and
allows the thermodynamic response of the disk to be controlled.

@@

We mostly focused on vertically isothermal disks ($\Gamma=1$), since
this is the relevant case for protoplanetary disks that are irradiated
\citep{chiang97}. It also permits a largely analytical discussion.  Our analysis is an
extension of the linear models considered in \cite{lubow93} by 
adding vertical shear and (rapid) thermal relaxation, but in the limit of low-frequency
perturbations in nearly-Keplerian disks. These approximations were
made for simplicity, but also motivated by recent numerical
simulations \citep{nelson13}.

For isothermal perturbations ($\beta=0$) we explicitly
demonstrated instability  associated with vertical shear, and that the 
VSI growth rate is limited by the maximum vertical shear within the 
vertical domain of interest. The fundamental VSI mode is characterized
by one node in the density perturbation at the midplane 
(i.e. anti-symmetric about $z=0$), which corresponds to a vertical
velocity perturbation  $\delta v_z\sim\mathrm{constant}$ near the
mid-plane with zero nodes. These have been referred to as
`corrugation modes' by \cite{nelson13}, and were found to dominate
their numerical simulations, as well as that of
\cite{stoll14}. Furthermore, when computing unstable modes for a
disk with a physical model for thermal relaxation, we also find
the fundamental mode to be dominant and robust. Such modes are thus
likely most relevant to  realistic disks.

For non-isothermal  perturbations with finite thermal relaxation ($\gamma>1$,
$\beta>0$) , the general dispersion relation (Eq. \ref{relax_disp}) is
quite complicated. However, for the 
fundamental VSI some simplification is  possible. In particular, we
quantified the critical thermal relaxation timescale
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below which small-scale
perturbations may be regarded as isothermal and the fundamental VSI
operates.  We expect $\beta<\beta_\mathrm{crit}$ to be a sufficient
condition for the VSI to operate. We have therefore provided a
quantitative measure for the`fast' thermal relaxation requirement for the
VSI.

@@

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks, specifically the Minimum Mass Solar Nebulae
(MMSN). We consider the disk model described in \cite{chiang10}, and
modeled the thermal relaxation timescale $\beta$ based on
dust opacity, which results in a $\beta$ that depends on both the
lengthscale of interest as well as the height from the midplane.  
We find the MMSN is unstable to the fundamental VSI at $10$s of AU,
with characteristic growth times of $\sim 30$ orbits. Although $\beta$
is not spatially constant as assumed in our analysis, we find that
evaluating $\beta < \beta_\mathrm{crit}$ at the disk
midplane gives a rough measure of the radii at which the VSI operates in the
MMSN.  This simple criteron may be a useful first step to 
estimate the relevance of the VSI in realistic astrophysical
disks without resorting to non-linear numerical simulations.

>>>>

We included a source term in the energy equation which relaxes the disk
temperature to its initial value on a timescale
$t_c=\beta\Omega_k^{-1}$. This approach to model thermal relaxation was also
employed in the global numerical simulations of \cite{nelson13}, and
allows the thermodynamic response of the disk to be controlled.
However, it is possible to model $\beta$ self-consistently by
relating it to thermal timescales expected in realistic disks.

@@

We mostly focused on vertically isothermal disks ($\Gamma=1$), since
this is the relevant case for protoplanetary disks that are irradiated
\citep{chiang97}. It also permits a largely analytical discussion.
Our analysis is an extension of the linear models considered in
\cite{lubow93} by  adding vertical shear and (rapid) thermal
relaxation, but in the limit of low-frequency perturbations in
nearly-Keplerian disks. These approximations were made for simplicity,
but also motivated by recent numerical simulations
\citep{nelson13}. We obtained explicit solutions for the VSI with
thermal relaxation in the thin disk limit.

% For isothermal perturbations ($\beta=0$) we explicitly
% demonstrated instability  associated with vertical shear, and that the 
% VSI growth rate is limited by the maximum vertical shear within the 
% vertical domain of interest. 
% For perturbations with finite thermal relaxation, the general
% dispersion relation (Eq. \ref{relax_disp}) is quite
% complicated. However, for the fundamental VSI some simplification is
% possible. In particular,

The fundamental VSI mode is characterized by a vertical velocity
perturbation with zero nodes and $\delta v_z\sim\mathrm{constant}$
near the mid-plane, and by one node in the density perturbation at the
midplane  (i.e. anti-symmetric about $z=0$).  These have been referred
to as `corrugation modes' by \cite{nelson13}, and were found to
dominate their numerical simulations, as well as that of
\cite{stoll14}. Furthermore, when computing unstable modes for a
protoplanetary disk model with thermal relaxation based on realistic
disk properties, we find the fundamental mode to be dominant and
robust against boundary conditions. The fundamental mode is thus
likely the most relevant VSI mode in protoplanetary disks.

We quantified the critical thermal relaxation timescale 
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below which small-scale
perturbations may be regarded as isothermal and the fundamental VSI
operates. %  We expect $\beta<\beta_\mathrm{crit}$ to be a sufficient
% condition for the VSI to operate.
We have therefore provided a quantitative measure for the `fast'
thermal relaxation requirement for the VSI in astrophysical disks.

@@

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks, specifically the Minimum Mass Solar Nebulae
(MMSN). We consider the disk model described in \cite{chiang10}, and
modeled the thermal relaxation timescale $\beta$ based on
dust opacity, which results in a $\beta$ that depends on both the
perturbation lengthscale as well as the height from the midplane.  
We find the MMSN is unstable to the fundamental VSI at $10$s of AU,
with characteristic growth times of $\sim 30$ orbits. Although $\beta$
is not spatially constant as assumed in our analysis, we find that
evaluating $\beta < \beta_\mathrm{crit}$ at the disk
midplane gives a rough measure of the radii at which the VSI operates in the
MMSN.  This simple criteron may be a useful first step to 
estimate the relevance of the VSI in realistic astrophysical
disks without resorting to non-linear numerical simulations.
